Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-10 RevisedSubmission/hines_hetland_2017_gji.tex
did more

Here we propose using Gaussian process regression (GPR) \citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR is closely related to kriging \citep{Cressie1992} and least squares collocation \citep{Moritz1978}. The latter has been used by \citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain rates from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered data. Since GNSS stations are irregularly spaced and observation times may differ between stations, GPR is an ideal tool for synthesizing discrete GNSS data into a spatially and temporally continuous representation of surface deformation. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a continuous stochastic process where any collection of points in the process can be described as a multivariate Gaussian random variable. Just as multivariate Gaussian random variables are fully determined by a mean vector and a covariance matrix, a Gaussian process is fully determined by a mean function and a covariance function. For example, Brownian motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has a mean function $\mathrm{E}[B(t)] = 0$ and a covariance function $\mathrm{Cov}\left[B(t),B(t')\right] = \min(t,t')$, where $t,t' \ge 0$. If there is no information available to help choose an appropriate prior Gaussian process, then maximum likelihood methods can be used to objectively choose one that is most consistent with the data. Once a prior is selected, we incorporate GNSS observations with the prior to form a posterior estimate of transient strain.  The posterior transient strain is also a Gaussian process, and we can use its distribution to confidently discern geophysical signal from noise. We use GPR to infer strain resulting from SSEs in the Pacific Northwest, demonstrating that GPR is an effective tool for detecting transient geophysical processes.

@@

\subsection{Outlier detection}\label{sec:Outlier}
In deriving our formulation for transient strain rates, we have assumed that noise in the data vector is normally distributed. This is not an appropriate assumption for GNSS data, which often have more outliers than would be predicted for normally distributed noise. Methods for analyzing GNSS data should either be robust against outliers or should involve a preprocessing step in which outliers are detected and removed. Examples of the former include the MIDAS method for estimating secular velocities \citep{Blewitt2016} and the GPS Imaging method for detecting spatially coherent features \citep{Hammond2016}.  In this study, we identify and remove outliers as a preprocessing step before estimating $\dot\varepsilon$. Outliers are identified based on the residuals for a model that best fits the observed data. Observations with residuals that exceed some threshold are removed. This strategy is commonly used to detect outliers in GNSS data, where the residuals are typically computed by estimating and removing a linear trend and seasonal terms for each time series \citep[e.g.,][]{Johansson2002,Dong2006,Bos2013}. The outlier detection method used here is different from this more standard approach because the model we fit to the data additionally consists of a spatially and temporally correlated Gaussian process, $v \sim \mathcal{N}(0,C_v)$. We use $v$ to explain any coherent features in the data that cannot be explained by the linear trend and sinusoids. With this more complex model, we are able to detect outliers that are spatially or temporally anomalous.

To describe our outlier detection algorithm in detail, we let $\mitbf{d}^*$ denote all $n$ GNSS displacement observations for a single component (note that we dropped the component subscripts since outliers are detected for the easting and northing components independently). Our goal is to find the index set of non-outliers in $\mitbf{d}^*$, which we denote as $\mitbf{\Omega}$. For notational ease, we use a tilde to indicate that an array only contains elements corresponding to $\mitbf{\Omega}$ (e.g., the vector of non-outlier observations is denoted $\tilde{\mitbf{d}}^* = [d_i^*]_{i \in \mitbf{\Omega}}$). We now consider $\mitbf{d}^*$ to be a realization of the random vector
\begin{equation}
\mitbf{d} = \mitbf{G}\mitbf{m} + v(\mitbf{P}) + \mitbf{w},
\end{equation}
where $\mitbf{G}$ and $\mitbf{m}$ are the same as in eq. (\ref{eq:DataVec}), and $\mitbf{w}$ is a vector of uncorrelated noise with assumed known standard deviations $\mitbf{\sigma} = [\sigma_1, \sigma_2, ..., \sigma_n]$. Our best fitting model for $\mitbf{d}^*$ is taken to be the mean of $\mitbf{G}\mitbf{m} + v(\mitbf{P})$ after it is conditioned with non-outlier observations. The residual vector is then
\begin{equation}\label{eq:Residual1}
\mitbf{r} = \mitbf{d}^* - \mathrm{E}\left[\Big(\mitbf{G}\mitbf{m} + v(\mitbf{P})\Big) \Big| \Big(\tilde{\mitbf{d}} = \tilde{\mitbf{d}}^*\Big)\right].
\end{equation}
Following a similar procedure as above, we can write eq. (\ref{eq:Residual1}) as
\begin{equation}\label{eq:Residual}
\mitbf{r} = \mitbf{d}^*  - 
            \left[\begin{array}{cc}
                  C_v(\mitbf{P},\tilde{\mitbf{P}}) & \mitbf{G} \\
                  \end{array}\right]
            \left[\begin{array}{cc}
                  C_v(\tilde{\mitbf{P}},\tilde{\mitbf{P}}) + \mathrm{diag}(\tilde{\mitbf{\sigma}}^2) & \tilde{\mitbf{G}} \\
                  \tilde{\mitbf{G}}^T  & \mitbf{0} \\
                  \end{array}\right]^{-1}
            \left[\begin{array}{c}
                  \tilde{\mitbf{d}}^* \\
                  \mitbf{0} \\
                  \end{array}\right].
\end{equation}
We start the algorithm by letting $\mitbf{\Omega}$ contain all $n$ indices, and we refine $\mitbf{\Omega}$ iteratively. For each iteration, we compute $\mitbf{r}$ with eq. (\ref{eq:Residual}) and then update $\mitbf{\Omega}$ so that it contains the indices of $\mitbf{r}$ whose weighted values are less than $\lambda$ times the weighted root mean square of $\tilde{\mitbf{r}}$, 
\begin{equation}\label{eq:Update}
\mitbf{\Omega} \leftarrow \left\{i : \left|\frac{r_i}{\sigma_i}\right| < \lambda \cdot \sqrt{\frac{1}{|\mitbf{\Omega}|} \sum_{j \in \mitbf{\Omega}} \frac{r_j^2}{\sigma_j^2}} \right\}.
\end{equation} 
Iterations continue until the new $\mitbf{\Omega}$ is the same as the previous $\mitbf{\Omega}$.

We use $\lambda=4$ in this study, which in our experience accurately identifies outliers without unnecessarily decimating the data. Iterations continue until the new $\mitbf{\Omega}$ is equal to the previous $\mitbf{\Omega}$.

It should be noted that this algorithm does not identify jumps in GNSS time series (cite Gazeaux 2013), which are another common issue. Some, but not all, jumps can be automatically removed by looking up the dates of equipment changes and earthquakes. However, it is still necessary to manually find and remove jumps of unknown origin. That being said, this outlier detection algorithm significantly reduces the effort needed to manually clean GNSS data.

@@

>>>>

Here we propose using Gaussian process regression (GPR) \citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR is closely related to kriging \citep{Cressie1992} and least squares collocation \citep{Moritz1978}. The latter has been used by \citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain rates from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered data. Since GNSS stations are irregularly spaced and observation times may differ between stations, GPR is an ideal tool for synthesizing discrete GNSS data into a spatially and temporally continuous representation of surface deformation. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a continuous stochastic process whose value at any collection of points can be described as a multivariate Gaussian random variable. Just as a multivariate Gaussian random variable is fully determined by a mean vector and a covariance matrix, a Gaussian process is fully determined by a mean function and a covariance function. For example, Brownian motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has a mean function $\mathrm{E}[B(t)] = 0$ and a covariance function $\mathrm{Cov}\left[B(t),B(t')\right] = \min(t,t')$, where $t,t' \ge 0$. If there is no information available to help choose an appropriate prior Gaussian process, then maximum likelihood methods can be used to objectively choose one that is most consistent with the data. Once a prior is selected, we incorporate GNSS observations with the prior to form a posterior estimate of transient strain.  The posterior transient strain is also a Gaussian process, and we can use its distribution to confidently discern geophysical signal from noise. We use GPR to infer strain resulting from SSEs in the Pacific Northwest, demonstrating that GPR is an effective tool for detecting transient geophysical processes.

@@

\subsection{Outlier detection}\label{sec:Outlier}
In deriving our formulation for transient strain rates, we have assumed that noise in the data vector is normally distributed. This is not an appropriate assumption for GNSS data, which often have more outliers than would be expected for normally distributed noise. Methods for analyzing GNSS data should either be robust against outliers or should involve a preprocessing step in which outliers are detected and removed. Examples of the former include the MIDAS method for estimating secular velocities \citep{Blewitt2016} and the GPS Imaging method for detecting spatially coherent features \citep{Hammond2016}. In this study, we identify and remove outliers as a preprocessing step before estimating $\dot\varepsilon$. Outliers are identified based on the residuals for a model that best fits the observed data. Observations with residuals that exceed some threshold are removed. This strategy for detecting outliers is commonly used for GNSS data, where the model being fit to the data typically consists of a linear trend and seasonal terms for each GNSS station \citep[e.g.,][]{Johansson2002,Dong2006,Bos2013}. To prevent transient geophysical signal from being erroneously identified as outliers, the model used in our outlier detection algorithm additionally consists of a temporally correlated Gaussian process. The details of our algorithm are given in Appendix A.

It should be noted that our algorithm does not identify jumps in GNSS time series, which are another common issue. Some, but not all, jumps can be automatically removed by looking up the dates of equipment changes and earthquakes \citep{Gazeaux2013}. However, it is still necessary to manually find and remove jumps of unknown origin.

@@

\appendix

\section{Outlier detection algorithm}
Our outlier detection algorithm is loosely based on the data editing algorithm from \citet{Acheson1975}. Let $\mitbf{d}^*$ denote all $n$ GNSS displacement observations for a single directional component, which have been made at positions and times $\mitbf{P}$.  We model $\mitbf{d}^*$ as a realization of the random vector 
\begin{equation}\label{eq:OutlierDataVec}
\mitbf{d} = \mitbf{G}\mitbf{m} + v(\mitbf{P}) + \mitbf{w},
\end{equation}
where $\mitbf{G}$ and $\mitbf{m}$ are the same as in eq. (\ref{eq:DataVec}), $v$ is a Gaussian process distributed as $\mathcal{GP}(0,C_v)$, and $\mitbf{w}$ is a vector of uncorrelated noise with known standard deviations $\mitbf{\sigma} = [\sigma_1, \sigma_2, ..., \sigma_n]$. The Gaussian process $v$ is intended to describe transient features in the data that cannot be explained by the linear trend or seasonal terms in $\mitbf{G}$. We let the temporal covariance of $v$ be a squared exponential, and we let $v$ be spatially uncorrelated so that,
\begin{equation}
C_v(p,p') = \phi^2\exp\left(\frac{-|t - t'|^2}{2\tau^2}\right) C_w(\vec{x},\vec{x}'),
\end{equation}
where 
\begin{equation}\label{eq:WhiteCov}
C_w(\vec{x},\vec{x}') = 
\begin{cases}
1, &\vec{x} = \vec{x}' \\
0, &\vec{x} \neq \vec{x}'
\end{cases}.
\end{equation}
The spatial covariance of $v$ has little effect on the detected outliers, and so we have assumed that $v$ is spatially uncorrelated for simplicity. Based on our experience, $v$ can reasonably describe most transient features in the data when we set $\phi = 1$ mm and $\tau = 10$ days.

Our goal is to find the index set of non-outliers in $\mitbf{d}^*$, which we denote as $\mitbf{\Omega}$. We use a tilde to indicate that an array only contains elements corresponding to $\mitbf{\Omega}$ (e.g., the vector of non-outlier observations is denoted $\tilde{\mitbf{d}}^* = [d_i^*]_{i \in \mitbf{\Omega}}$). The outliers are identified iteratively, and we initiate $\mitbf{\Omega}$ with all $n$ indices. We consider outliers to be data that are inconsistent with eq. (\ref{eq:OutlierDataVec}), which we determine with the residual vector 
\begin{align}\label{eq:Residual}
\mitbf{r} &= \mitbf{d}^* - \mathrm{E}\left[\Big(\mitbf{G}\mitbf{m} + v(\mitbf{P})\Big) \Big| \Big(\tilde{\mitbf{d}} = \tilde{\mitbf{d}}^*\Big)\right] \\
          &= \mitbf{d}^*  - 
            \left[\begin{array}{cc}
                  C_v(\mitbf{P},\tilde{\mitbf{P}}) & \mitbf{G} \nonumber \\
                  \end{array}\right]
            \left[\begin{array}{cc}
                  C_v(\tilde{\mitbf{P}},\tilde{\mitbf{P}}) + \mathrm{diag}(\tilde{\mitbf{\sigma}}^2) & \tilde{\mitbf{G}} \\
                  \tilde{\mitbf{G}}^T  & \mitbf{0} \\
                  \end{array}\right]^{-1}
            \left[\begin{array}{c}
                  \tilde{\mitbf{d}}^* \\
                  \mitbf{0} \\
                  \end{array}\right].
\end{align}
Data with abnormally large residuals are identified as outliers. For each iteration, we compute $\mitbf{r}$ and then update $\mitbf{\Omega}$ so that it contains the indices of $\mitbf{r}$ whose weighted values are less than $\lambda$ times the weighted root mean square of $\tilde{\mitbf{r}}$, 
\begin{equation}\label{eq:Update}
\mitbf{\Omega} \leftarrow \left\{i : \left|\frac{r_i}{\sigma_i}\right| < \lambda \cdot \sqrt{\frac{1}{|\mitbf{\Omega}|} \sum_{j \in \mitbf{\Omega}} \frac{r_j^2}{\sigma_j^2}} \right\}.
\end{equation} 
Iterations continue until the new $\mitbf{\Omega}$ is the same as the previous $\mitbf{\Omega}$. We use $\lambda=4$ in this study, which in our experience accurately identifies outliers without unnecessarily decimating the data.

To demostrate this algorithm ...
